Viscous Transport in Eroding Porous Media

### 2019-09-24 docs/transport/erosion_transport.tex
edited up to numerical results

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using the Barycentric quadrature rule, the velocity, shear stress,
and vorticity are computed with an error that is bounded independent of
the target location. However, applied directly, it requires
$\mathcal{O}(N^2)$ operations.  By using a fast summation method, such
as the fast multipole method~\cite{gre-rok1987}, this cost can be
reduced to $\mathcal{O}(N)$ operations.  However, the Barycentric
quadrature rules require several $N$-body calculations, and they must be
done for each component of $\bd\Omega$ individually.  The resulting cost
would be prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and the trapezoid rule.  In short, the fast
multipole accelerated trapezoid rule is used to compute the velocity at
all the discretization points, and the non-accelerated Barycentric
quadrature rule is used to locally correct the errors introduced by the
trapezoid rule.  Note that the source points of the layer potential is
always one of the eroding bodies or the outer wall, but the target point
can either be on another component of $\bd\Omega$ or it can be in the
fluid bulk $\Omega$.

\todo[inline]{left off here}
To compute the velocity double-layer potential, we start by applying the
trapezoid rule~\eqref{eqn:trap} accelerated with the FMM.  This
calculation requires $\mathcal{O}(N)$ operations, and we call the
resulting velocity $\vv_\trap(\xx)$.  Depending on the number of
discretization points $N_\iin$ and $N_\out$, there is a region separated
from $\bd\Omega$ where the error of $\vv_\trap(\xx)$ is less than a
user-specified tolerance.  However, for the other points in $\Omega$,
the trapezoid rule needs to be replaced with a more accurate quadrature
rule.  Note, however, that since a point is typically only close to one
or two components of $\bd\Omega$, only the contribution of these nearby
bodies need to be replaced.  Assuming that $\xx$ is too close to
$\gamma_k$, we first subtract the inaccurate approximation of the
double-layer potential due to $\gamma_k$.  Then, the Barycentric
quadrature rule is used to compute the velocity due to $\gamma_k$ with
more accuracy.  That is, the final velocity is
\begin{align}
  \vv(\xx) = \vv_\trap(\xx) - \frac{L_k}{N_k} \sum_{j=1}^{N_\iin} 
    D(\xx,\yy^k_j) \eeta(\yy_j) + \vv^k_\bary(\xx),
  \label{eqn:velocityDecomp}
\end{align}
where $\vv^k_\bary(\xx)$ is the velocity at $\xx$ resulting from
applying the Barycentric quadrature rule to the double-layer potential
due to $\gamma_k$.  This strategy naturally extends to points that are
close to $\Gamma$, and to the scenario when $\xx$ is simultaneously
close to multiple components of $\bd\Omega$.  While the term
$\vv_\trap(\xx)$ in equation~\eqref{eqn:velocityDecomp} is computed for
all target points using the FMM, the other two terms are computed with a
direct summation.  However, these terms are only required for target
points near an eroding body or the outer wall, and these points make up
only a small fraction of the total number of points.

An identical strategy is used to compute the vorticity and the
deformation tensor.  That is, the trapezoid rule is used as a first pass
to form the vorticity and deformation tensor, and then local corrections
are made to replace the inaccuracies of the trapezoid rule.  However,
since the shear stress and vorticity are only computed once per time
step, the trapezoid rule is applied with a direct summation rather than
the FMM.  Relative to the cost of computing the velocity at each
iteration of GMRES with the FMM, the additional once-per-time-step cost
to compute the vorticity and deformation tensor is minimal.

Per grid point, applying the Barycentric quadrature rules dominates the
computational cost. Therefore, it is imperative that we only apply the
Barycentric quadrature when the trapezoid rule is inaccurate.  As a rule
of thumb, the trapezoid rule due to $\gamma_k$ achieves machine epsilon
accuracy if~\cite{bar2014}
\begin{align}
  \label{eqn:TrapCutoff}
  d(\xx,\gamma_k) = \inf_{\yy \in \gamma_k} \|\xx - \yy\| > 
    5 \frac{L_k}{N_\iin}.
\end{align}
To avoid checking if all target point $\xx$
satisfy~\eqref{eqn:TrapCutoff}, we make use of the {\thL}
variables~\cite{hou-low-she1994}.  First, given two eroding bodies $\gamma_i$ and $\gamma_j$ with
lengths $L_i$ and $L_j$, and centers $\cc_i$ and $\cc_j$, we check if
\begin{align}
  \label{eqn:BodyCutoff}
  \|\cc_i - \cc_j\| < \frac{L_i}{2\pi} + \frac{L_j}{2\pi} + 
    \alpha_\iin \left(\frac{L_i}{N_\iin} + \frac{L_j}{N_\iin} \right),
\end{align}
where $\alpha_\iin \geq 1$ is a parameter that needs to be determined.
In this manner, rather than using an expensive all-to-all algorithm to
compute the distance between all pairs of discretization points, we
compute the distance between pairs of circle centers.  This criteria
allows us to quickly determine bodies that contain discretization points
where the Barycentric quadrature rule may need to be applied, and the
parameter $\alpha_\iin$ accounts for the fact that the bodies are being
approximated with circles.  Assuming that the two bodies $\gamma_i$ and
$\gamma_j$ satisfy condition~\eqref{eqn:BodyCutoff}, for each point $\xx
\in \gamma_j$, we check if
\begin{align}
  \label{eqn:PointsBodyCutoff}
  \|\xx - \cc_i\| < \frac{L_i}{2\pi}
+ \alpha_\iin \left(\frac{L_i}{N_\iin} \right).
\end{align}
To determine if points on $\gamma_i$ are too close to the outer wall, we
recall that the geometry is contained in a rounded-off version of
$[-3,3] \times [-1,1]$, and that the bodies are all contained in $[-1,1]
\times [-1,1]$, so a target point can only be close to the lines $y =
\pm 1$. Therefore, we first check if
\begin{align}
  \left\|\cc_i - \left[
    \begin{array}{c}
      x_i \\ \pm 1
    \end{array}
    \right]
  \right\| < \frac{L_i}{2 \pi} + \alpha_\out \frac{L_\out}{N_\out},
\end{align}
where $x_i$ is the $x$-coordinate of $\cc_i$. If body $\gamma_i$
satisfies this condition, for each point $\xx=(x,y) \in \gamma_i$, we
apply the Barycentric rule to points that satisfy
\begin{align}
  \label{eqn:PointsWallCutoff}
  |y \pm 1| < \frac{L_i}{2 \pi} +\alpha_\out \frac{L_\out}{N_\out}.
\end{align}
Finally, to determine if a target point $\xx$ in the fluid bulk requires
the Barycentric quadrature rule, we only check
conditions~\eqref{eqn:PointsBodyCutoff}
and~\eqref{eqn:PointsWallCutoff}.

@@

Rather than tracking the $(x,y)$ coordinates, the {\thL} variables are
tracked. In addition, a tangential velocity field is added to each body
so that the discretization points remain equispaced in arclength for all
time.  A second-order Implicit-Explicit Runge-Kutta method is used. In
particular, the diffusive term corresponding to the curvature
penalization term is discretized implicitly, and all other terms, which
are non-stiff, are treated explicitly.  By using this time stepping
method in conjunction with the Barycentric rule, we stably evolve the
eroding bodies.  Finally, to increase the overall stability of the
method, a narrow Gaussian filter is applied to the erosion rate at each
time step.

@@

Once a collection of trajectories are formed, they are used to quantify
the dispersion and the tortuosity.  The dispersion simply computes the
second-moment of the trajectories as described in
equation~\eqref{eqn:moments}.  We use $N_p = 1000$ streamlines so that
the statistics have converged~\cite{bel-sal-rin1992}.  However, as
described in Section~\ref{sec:dispersion}, a reinsertion method is used
to compute trajectories that are sufficiently long to observe an
asymptotic anomalous dispersion rate.  To compute the tortuosity using
the trajectories, we consider trajectories crossing between the two
cross-sections $x=-1$ and $x=1$.  Then, we discretize
equation~\eqref{eqn:tortuosity1} as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_1(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_1(y_i) \Delta y}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this section, we apply our new quadrature method to (1) simulate
erosion of several different configurations of eroding bodies, and (2)
characterize transport inside eroded geometries.  Each eroding body is
initially a circle with center $\cc_i$ and radius $r_i$.  The center and
radius are chosen at random, and the body is accepted as long as it is
contained in $[-1,1] \times [-1,1]$ and is sufficiently separated from
all other bodies.  Since the error of our quadrature method is bounded
independent of the geometry, we consider bodies that are separated by
less than 10\% of an arclength spacing.  The randomized method is
repeated until the initial geometry reaches a desired initial porosity.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using the Barycentric quadrature rule, the velocity, shear stress,
and vorticity are computed with an error that is bounded independent of
the target location. However, applied directly, it requires
$\mathcal{O}(N^2)$ operations.  By using a fast summation method, such
as the fast multipole method~\cite{gre-rok1987}, this cost can be
reduced to $\mathcal{O}(N)$ operations.  However, the Barycentric
quadrature rules require several $N$-body calculations, and they must be
done for each component of $\bd\Omega$ individually.  The resulting cost
would be prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and the trapezoid rule.  Note that the
source points of the layer potential is always one of the eroding bodies
or the outer wall, but the target point can either be on another
component of $\bd\Omega$ or it can be in the fluid bulk $\Omega$.

To compute the velocity double-layer potential, we start by applying the
trapezoid rule~\eqref{eqn:trap} accelerated with the FMM.  This
calculation requires $\mathcal{O}(N)$ operations, and we call the
resulting velocity $\vv_\trap(\xx)$.  Since the trapezoid rule is
spectrally accurate, the error of $\vv_\trap(\xx)$ is small for points
sufficiently far from $\bd\Omega$, and this region depends on the number
of discretization points $N_\iin$ and $N_\out$.  However, the trapezoid
rule needs to be replaced with a more accurate quadrature rule for
points that are too close to $\bd\Omega$.  Note, however, that since a
point is typically only close to one or two components of $\bd\Omega$,
only the contribution of these nearby bodies needs to be replaced.
Assuming that $\xx$ is too close to $\gamma_k$, we first subtract the
inaccurate trapezoid rule approximation of the double-layer potential
due to $\gamma_k$.  Then, the Barycentric quadrature rule is used to
compute the velocity due to $\gamma_k$ with more accuracy.  That is, the
final velocity is
\begin{align}
  \vv(\xx) = \vv_\trap(\xx) - \frac{L_k}{N_k} \sum_{j=1}^{N_\iin} 
    D(\xx,\yy^k_j) \eeta(\yy_j) + \vv^k_\bary(\xx),
  \label{eqn:velocityDecomp}
\end{align}
where $\vv^k_\bary(\xx)$ is the velocity at $\xx$ resulting from
applying the Barycentric quadrature rule to the double-layer potential
due to $\gamma_k$.  This strategy naturally extends to points that are
close to $\Gamma$, and to the scenario when $\xx$ is simultaneously
close to multiple components of $\bd\Omega$.  While the term
$\vv_\trap(\xx)$ in equation~\eqref{eqn:velocityDecomp} is computed for
all target points using the FMM, the other two terms are computed with a
direct summation.  However, these terms are only required for target
points near an eroding body or the outer wall, and these points make up
only a small fraction of the total number of points.

An identical strategy is used to compute the vorticity and the
deformation tensor.  That is, the trapezoid rule is used as a first pass
to form the vorticity and deformation tensor, and then local corrections
are made to replace the inaccuracies of the trapezoid rule.  However,
since the shear stress and vorticity are only computed once per time
step, the trapezoid rule is applied with a direct summation rather than
the FMM.  Relative to the cost of computing the velocity at each
iteration of GMRES with the FMM, the additional once-per-time-step costs
to compute the vorticity and deformation tensor are minimal.

Per grid point, applying the Barycentric quadrature rules dominates the
computational cost. Therefore, it is imperative that we only apply the
Barycentric quadrature when the trapezoid rule is inaccurate.  As a rule
of thumb, the trapezoid rule due to $\gamma_k$ achieves machine epsilon
accuracy if~\cite{bar2014}
\begin{align}
  \label{eqn:TrapCutoff}
  d(\xx,\gamma_k) = \inf_{\yy \in \gamma_k} \|\xx - \yy\| > 
    5 \frac{L_k}{N_\iin}.
\end{align}
To avoid checking if all target point $\xx$
satisfy~\eqref{eqn:TrapCutoff}, we make use of the {\thL}
variables~\cite{hou-low-she1994}.  First, given two eroding bodies $\gamma_i$ and $\gamma_j$ with
lengths $L_i$ and $L_j$, and centers $\cc_i$ and $\cc_j$, we check if
\begin{align}
  \label{eqn:BodyCutoff}
  \|\cc_i - \cc_j\| < \frac{L_i}{2\pi} + \frac{L_j}{2\pi} + 
    \alpha_\iin \left(\frac{L_i}{N_\iin} + \frac{L_j}{N_\iin} \right),
\end{align}
where $\alpha_\iin \geq 1$ is a parameter that needs to be determined.
In this manner, rather than using an expensive all-to-all algorithm to
compute the distance between all pairs of discretization points, we
compute the distance between pairs of circle centers.  This criteria
allows us to quickly determine bodies that contain discretization points
where the Barycentric quadrature rule might need to be applied, and the
parameter $\alpha_\iin$ accounts for the fact that the bodies are being
approximated with circles.  Assuming that the two bodies $\gamma_i$ and
$\gamma_j$ satisfy condition~\eqref{eqn:BodyCutoff}, for each point $\xx
\in \gamma_j$, we check if
\begin{align}
  \label{eqn:PointsBodyCutoff}
  \|\xx - \cc_i\| < \frac{L_i}{2\pi}
+ \alpha_\iin \left(\frac{L_i}{N_\iin} \right).
\end{align}
To determine if points on $\gamma_i$ are too close to the outer wall, we
recall that the eroding bodies are all contained in $[-1,1] \times
[-1,1]$, so a target point can only be close to the lines $y = \pm 1$.
Therefore, we first check if
\begin{align}
  \left\|\cc_i - \left[
    \begin{array}{c}
      x_i \\ \pm 1
    \end{array}
    \right]
  \right\| < \frac{L_i}{2 \pi} + \alpha_\out \frac{L_\out}{N_\out},
\end{align}
where $x_i$ is the $x$-coordinate of $\cc_i$. If body $\gamma_i$
satisfies this condition, for each point $\xx=(x,y) \in \gamma_i$, we
apply the Barycentric rule to points that satisfy
\begin{align}
  \label{eqn:PointsWallCutoff}
  |y \pm 1| < \frac{L_i}{2 \pi} +\alpha_\out \frac{L_\out}{N_\out}.
\end{align}
Finally, to determine if a target point $\xx$ in the fluid bulk requires
the Barycentric quadrature rule, we only check
conditions~\eqref{eqn:PointsBodyCutoff}
and~\eqref{eqn:PointsWallCutoff}.

@@

Rather than tracking the $(x,y)$ coordinates, the {\thL} variables are
tracked. In addition, a tangential velocity field is added to each body
so that the discretization points remain equispaced in arclength for all
time.  Additionally, to increase the overall stability of the method, a
narrow Gaussian filter is applied to the erosion rate at each time step.
Time stepping is performed with a second-order Implicit-Explicit
Runge-Kutta method. In particular, the diffusive term corresponding to
the curvature penalization term is discretized implicitly, and all other
terms, which are non-stiff, are treated explicitly.  By using this time
stepping method in conjunction with the Barycentric rule, we stably
evolve the eroding bodies.

@@

Once a collection of trajectories are formed, they are used to quantify
the dispersion and the tortuosity.  We use $N_p = 1000$ streamlines so
that the statistics have converged~\cite{bel-sal-rin1992}.  However, as
described in Section~\ref{sec:dispersion}, a reinsertion method is used
to compute trajectories that are sufficiently long to observe an
asymptotic anomalous dispersion rate.  To compute the tortuosity using
the trajectories, we consider trajectories crossing between the two
cross-sections $x=-1$ and $x=1$.  Then, we discretize
equation~\eqref{eqn:tortuosity1} as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_1(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_1(y_i) \Delta y}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{here}
In this section, we apply our new quadrature method to (1) simulate
erosion of several different configurations of eroding bodies, and (2)
characterize transport inside eroded geometries.  Each eroding body is
initially a circle with center $\cc_i$ and radius $r_i$.  The center and
radius are chosen at random, and the body is accepted as long as it is
contained in $[-1,1] \times [-1,1]$ and is sufficiently separated from
all other bodies.  Since the error of our quadrature method is bounded
independent of the geometry, we consider bodies that are separated by
less than 10\% of an arclength spacing.  The randomized method is
repeated until the initial geometry reaches a desired initial porosity.
